---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

# bidser

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/bidser)](https://CRAN.R-project.org/package=bidser)
[![R-CMD-check](https://github.com/bbuchsbaum/bidser/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/bbuchsbaum/bidser/actions/workflows/R-CMD-check.yaml)
[![pkgdown](https://github.com/bbuchsbaum/bidser/actions/workflows/pkgdown.yaml/badge.svg)](https://bbuchsbaum.github.io/bidser/)
[![Codecov test coverage](https://codecov.io/gh/bbuchsbaum/bidser/branch/master/graph/badge.svg)](https://app.codecov.io/gh/bbuchsbaum/bidser?branch=master)
<!-- badges: end -->

**bidser** reads and queries [BIDS](https://bids.neuroimaging.io/)
(Brain Imaging Data Structure) neuroimaging projects in R: locate subjects,
sessions, tasks, and files; resolve sidecar metadata; discover derivative
pipelines; and extract fMRIPrep confounds.

[Documentation](https://bbuchsbaum.github.io/bidser/) ·
[Getting started](https://bbuchsbaum.github.io/bidser/articles/quickstart.html) ·
[Derivatives](https://bbuchsbaum.github.io/bidser/articles/derivatives.html) ·
[Confounds](https://bbuchsbaum.github.io/bidser/articles/confounds-and-variables.html) ·
[API reference](https://bbuchsbaum.github.io/bidser/reference/) ·
[Changelog](NEWS.md)

## Installation

Install the released version from CRAN:

```r
install.packages("bidser")
```

Or the development version from GitHub:

```r
# install.packages("remotes")
remotes::install_github("bbuchsbaum/bidser")
```

CRAN currently publishes 0.5.0; this repository is at 0.5.1.

## Quick start

Load a BIDS project and query functional scans with the public entry points
`bids_project()`, `participants()`, `tasks()`, and `func_scans()`:

```{r}
library(bidser)

ds_path <- get_example_bids_dataset("ds001")
proj <- bids_project(ds_path)

participants(proj)
tasks(proj)
head(func_scans(proj, full_path = FALSE))
```

Related APIs include `read_events()`, `query_files()`, `get_metadata()`,
`derivative_pipelines()`, and `create_mock_bids()` for offline fixtures.
See the vignettes linked above for worked examples.

## fMRIPrep confounds

`read_confounds()` selects nuisance regressors from fMRIPrep confound
tables. Prefer the public, version-robust helpers over hand-listed column
names:

```r
# Named, version-robust sets (resolve to whatever columns your dataset has)
read_confounds(proj, cvars = confound_set("motion24"))
read_confounds(proj, cvars = confound_set("36p"))

# PCA + raw denoising strategies (recommended modern default)
read_confounds(proj, cvars = confound_strategy("pcabasic80"))

# Discover what is available
list_confound_sets()
list_confound_strategies()
```

Code that previously reached into the unexported
`bidser:::DEFAULT_CVARS2` should switch to the stable public handle
`confound_set("legacy_default")`, which returns the identical 26-name set.
See `?read_confounds`, `?confound_set`, and the
[confounds vignette](https://bbuchsbaum.github.io/bidser/articles/confounds-and-variables.html).
