---
title: Working with Confounds and Variables
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{Working with Confounds and Variables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```



```{r load-packages}
library(bidser)
library(dplyr)
library(tidyr)
library(tibble)
```

After loading a BIDS dataset and finding your scans, the next step in most
fMRI workflows is extracting **confound regressors** and **event tables** for
modeling. bidser gives you tidy tibbles at every level -- per-run, per-subject,
or dataset-wide -- so you can go straight from BIDS into `lm()`, `lme4`, or
whatever modeling framework you prefer.

## Building a test dataset

We'll create a small mock dataset with two subjects, two tasks, and realistic
fMRIPrep-style confound files so that every code chunk in this vignette runs
without network access.

```{r build-mock, include = FALSE}
temp_dir <- tempfile("bids_confounds_vignette_")
dir.create(temp_dir)

# participants.tsv
readr::write_tsv(
  tibble(participant_id = c("sub-01", "sub-02"), age = c(25L, 30L), sex = c("M", "F")),
  file.path(temp_dir, "participants.tsv")
)

# dataset_description.json
jsonlite::write_json(
  list(Name = "ConfoundDemo", BIDSVersion = "1.7.0"),
  file.path(temp_dir, "dataset_description.json"),
  auto_unbox = TRUE
)

# Create raw + derivative structure
set.seed(42)
n_vols <- 100

for (sub in c("01", "02")) {
  for (task in c("rest", "nback")) {
    # --- raw ---
    func_dir <- file.path(temp_dir, paste0("sub-", sub), "func")
    dir.create(func_dir, recursive = TRUE, showWarnings = FALSE)

    bold_name <- sprintf("sub-%s_task-%s_run-01_bold.nii.gz", sub, task)
    file.create(file.path(func_dir, bold_name))

    events_name <- sprintf("sub-%s_task-%s_run-01_events.tsv", sub, task)
    n_trials <- if (task == "nback") 40 else 0
    if (n_trials > 0) {
      ev <- tibble(
        onset = sort(runif(n_trials, 0, 290)),
        duration = rep(2, n_trials),
        trial_type = sample(c("0back", "2back"), n_trials, replace = TRUE),
        response_time = round(rnorm(n_trials, 0.8, 0.15), 3)
      )
      readr::write_tsv(ev, file.path(func_dir, events_name))
    } else {
      readr::write_tsv(
        tibble(onset = numeric(0), duration = numeric(0), trial_type = character(0)),
        file.path(func_dir, events_name)
      )
    }

    # --- derivatives/fmriprep ---
    deriv_func <- file.path(temp_dir, "derivatives", "fmriprep",
                            paste0("sub-", sub), "func")
    dir.create(deriv_func, recursive = TRUE, showWarnings = FALSE)

    preproc_name <- sprintf(
      "sub-%s_task-%s_run-01_space-MNI152NLin2009cAsym_desc-preproc_bold.nii.gz",
      sub, task
    )
    file.create(file.path(deriv_func, preproc_name))

    conf_name <- sprintf(
      "sub-%s_task-%s_run-01_desc-confounds_timeseries.tsv",
      sub, task
    )
    conf <- tibble(
      csf = rnorm(n_vols), white_matter = rnorm(n_vols), global_signal = rnorm(n_vols),
      framewise_displacement = abs(rnorm(n_vols, 0.2, 0.1)),
      trans_x = cumsum(rnorm(n_vols, 0, 0.02)),
      trans_y = cumsum(rnorm(n_vols, 0, 0.02)),
      trans_z = cumsum(rnorm(n_vols, 0, 0.02)),
      rot_x   = cumsum(rnorm(n_vols, 0, 0.001)),
      rot_y   = cumsum(rnorm(n_vols, 0, 0.001)),
      rot_z   = cumsum(rnorm(n_vols, 0, 0.001)),
      trans_x_derivative1 = c(NA, diff(cumsum(rnorm(n_vols, 0, 0.02)))),
      trans_y_derivative1 = c(NA, diff(cumsum(rnorm(n_vols, 0, 0.02)))),
      trans_z_derivative1 = c(NA, diff(cumsum(rnorm(n_vols, 0, 0.02)))),
      rot_x_derivative1   = c(NA, diff(cumsum(rnorm(n_vols, 0, 0.001)))),
      rot_y_derivative1   = c(NA, diff(cumsum(rnorm(n_vols, 0, 0.001)))),
      rot_z_derivative1   = c(NA, diff(cumsum(rnorm(n_vols, 0, 0.001)))),
      trans_x_power2 = trans_x^2, trans_y_power2 = trans_y^2, trans_z_power2 = trans_z^2,
      rot_x_power2 = rot_x^2, rot_y_power2 = rot_y^2, rot_z_power2 = rot_z^2,
      trans_x_derivative1_power2 = trans_x_derivative1^2,
      trans_y_derivative1_power2 = trans_y_derivative1^2,
      trans_z_derivative1_power2 = trans_z_derivative1^2,
      rot_x_derivative1_power2 = rot_x_derivative1^2,
      rot_y_derivative1_power2 = rot_y_derivative1^2,
      rot_z_derivative1_power2 = rot_z_derivative1^2,
      a_comp_cor_00 = rnorm(n_vols), a_comp_cor_01 = rnorm(n_vols),
      a_comp_cor_02 = rnorm(n_vols), a_comp_cor_03 = rnorm(n_vols),
      a_comp_cor_04 = rnorm(n_vols), a_comp_cor_05 = rnorm(n_vols),
      t_comp_cor_00 = rnorm(n_vols), t_comp_cor_01 = rnorm(n_vols),
      t_comp_cor_02 = rnorm(n_vols),
      cosine_00 = cos(seq(0, pi, length.out = n_vols)),
      cosine_01 = cos(seq(0, 2 * pi, length.out = n_vols)),
      cosine_02 = cos(seq(0, 3 * pi, length.out = n_vols))
    )
    readr::write_tsv(conf, file.path(deriv_func, conf_name))
  }
}

# Also write dataset_description.json for fmriprep derivative
deriv_root <- file.path(temp_dir, "derivatives", "fmriprep")
jsonlite::write_json(
  list(Name = "fMRIPrep", BIDSVersion = "1.7.0",
       GeneratedBy = list(list(Name = "fMRIPrep", Version = "21.0.0"))),
  file.path(deriv_root, "dataset_description.json"),
  auto_unbox = TRUE
)
```

```{r load-project}
proj <- bids_project(temp_dir, fmriprep = TRUE)
proj
```

Our demo has `r length(participants(proj))` subjects, each with a resting-state
and an n-back task run, plus fMRIPrep derivatives containing confound
timeseries.

## Confound sets: predefined recipes

Before reading any files, it helps to know what confound sets bidser ships
with. These are curated variable lists that match common denoising strategies
from the fMRI literature.

```{r list-sets}
list_confound_sets()
```

Each set is a named collection of confound variable names. You can inspect
what a set contains:

```{r show-motion-sets}
# 6 rigid-body motion parameters
confound_set("motion6")

# Friston 24-parameter expansion (motion + derivatives + squares)
length(confound_set("motion24"))
```

The sets compose naturally -- `"36p"` is `motion24` plus tissue signals and
their expansions, while `"9p"` is `motion6` plus the three global signals.

## Reading confounds

`read_confounds()` reads fMRIPrep confound TSV files and returns tidy tibbles.
Pass a confound set name to select variables:

```{r read-motion6}
conf_nested <- read_confounds(proj, cvars = confound_set("motion6"))
conf_nested
```

The result is a nested tibble -- one row per run, with the actual timeseries
tucked inside the `data` list column. This structure makes it easy to
iterate over runs in a modeling pipeline:

```{r unnest-motion}
conf_nested |>
  unnest(data) |>
  select(participant_id, task, run, trans_x, rot_x) |>
  head()
```

### Selecting variables by name or wildcard

You can also pass a character vector of specific variable names, or use
wildcards to match patterns:

```{r wildcard-confounds}
# All CompCor components
compcor_conf <- read_confounds(
  proj,
  subid = "01", task = "rest",
  cvars = c("a_comp_cor_*", "t_comp_cor_*")
)
names(compcor_conf$data[[1]])
```

### Flat output for quick modeling

When you just want a single table for modeling, use `nest = FALSE`:

```{r flat-confounds}
conf_flat <- read_confounds(
  proj,
  cvars = confound_set("motion6"),
  nest = FALSE
)
dim(conf_flat)
head(conf_flat)
```

Each row is one volume, with subject/task/run identifiers alongside the
confound values -- ready to join with your design matrix.

## Confound strategies: PCA reduction

For high-dimensional confound sets, you often want to reduce the
dimensionality rather than include all regressors directly. A
`confound_strategy()` splits variables into two groups: those reduced via PCA
and those kept as-is.

```{r show-strategies}
list_confound_strategies()
```

The built-in `"pcabasic80"` strategy applies PCA to motion and CompCor
variables (retaining 80% of variance) while keeping cosine regressors
unchanged:

```{r pca-strategy}
strat <- confound_strategy("pcabasic80")
conf_pca <- read_confounds(proj, subid = "01", task = "rest", cvars = strat)
names(conf_pca$data[[1]])
```

The motion and CompCor variables have been replaced by a handful of principal
components, while the cosine regressors pass through untouched.

### Custom strategies

You can build your own strategy by specifying which variables get PCA-reduced
and which are kept raw:

```{r custom-strategy}
my_strat <- confound_strategy(
  pca_vars = c(confound_set("motion24"), confound_set("acompcor")),
  raw_vars = c("framewise_displacement", confound_set("cosine")),
  npcs = 5
)

conf_custom <- read_confounds(proj, subid = "01", task = "nback", cvars = my_strat)
names(conf_custom$data[[1]])
```

Five PCs from the motion + aCompCor space, plus FD and cosine regressors kept
in their original form.

## Reading event files

Event files describe the experimental design -- trial onsets, durations, and
conditions. `read_events()` returns them as nested tibbles:

```{r read-events}
events <- read_events(proj, task = "nback")
events
```

Unnest to get a flat trial table:

```{r unnest-events}
trials <- events |>
  unnest(data) |>
  select(.subid, .task, .run, onset, duration, trial_type, response_time)

head(trials)
```

```{r trial-summary}
trials |>
  group_by(.subid, trial_type) |>
  summarise(
    n_trials = n(),
    mean_rt = mean(response_time, na.rm = TRUE),
    .groups = "drop"
  )
```

If you want everything in one flat table directly, use `load_all_events()`:

```{r load-all-events}
all_events <- load_all_events(proj, task = "nback")
nrow(all_events)
```

## The variables table: one tibble per run

`variables_table()` pulls together scans, events, and confounds into a single
run-level tibble. This is the tidy bridge between BIDS and your analysis code:

```{r variables-table}
vars <- variables_table(proj)
vars |> select(.subid, .task, .run, any_of(c("n_scans", "n_events", "n_confound_rows")))
```

Each row is one run. The `scans`, `events`, and `confounds` list columns hold
the nested data. You can selectively include only events or only confounds:

```{r variables-events-only}
vars_events <- variables_table(proj, task = "nback", include = "events")
vars_events |> select(.subid, .task, .run, n_events)
```

This structure makes it straightforward to write a per-run analysis loop:

```{r per-run-loop}
# Use the nback-only table which has both events and confounds columns
vars_nback <- variables_table(proj, task = "nback")

vars_nback |>
  filter(n_events > 0, n_confound_rows > 0) |>
  rowwise() |>
  mutate(
    n_conditions = length(unique(events$trial_type)),
    mean_fd = mean(confounds$framewise_displacement, na.rm = TRUE)
  ) |>
  ungroup() |>
  select(.subid, .task, n_conditions, mean_fd)
```

## Dataset QA with bids_report()

`bids_report()` assembles a lightweight QA summary covering project metadata,
compliance checks, pipeline discovery, and run-level coverage:

```{r bids-report}
report <- bids_report(proj)
report
```

The underlying data is fully accessible for custom reporting:

```{r report-data}
rdata <- bids_report_data(proj)
rdata$summary
rdata$run_coverage
```

You can use the coverage table to spot missing data at a glance:

```{r coverage-check}
rdata$run_coverage |>
  mutate(
    has_scans = n_scans > 0,
    has_events = if ("n_events" %in% names(rdata$run_coverage)) n_events > 0 else FALSE,
    has_confounds = n_confound_rows > 0
  )
```

## Putting it together

A typical analysis script combines these pieces into a pipeline. Here is a
sketch that extracts design-ready data for each run:

```{r full-pipeline}
analysis_data <- variables_table(proj, task = "nback") |>
  filter(n_events > 0, n_confound_rows > 0) |>
  rowwise() |>
  mutate(
    n_trials = nrow(events),
    n_vols = nrow(confounds)
  ) |>
  ungroup() |>
  select(.subid, .task, .run, n_trials, n_vols)

analysis_data
```

From here you would unnest the events and confounds columns, build your design
matrix with `model.matrix()`, and fit your model -- all in native R tibbles
with no intermediate file-path bookkeeping.

```{r cleanup, include = FALSE}
unlink(temp_dir, recursive = TRUE)
```
