---
title: Work With Derivatives and Confounds
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header:
      - |-
        <script src="albers.js"></script>
        <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>
      - albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{Work With Derivatives and Confounds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

```{r theme-setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
}

suppressPackageStartupMessages({
  library(bidser)
  library(dplyr)
  library(tidyr)
})
```

```{r build-fixture, include = FALSE}
deriv_fixture <- tempfile("bidser-deriv-")
dir.create(deriv_fixture, recursive = TRUE)

readr::write_tsv(
  tibble::tibble(participant_id = "sub-01"),
  file.path(deriv_fixture, "participants.tsv")
)

jsonlite::write_json(
  list(Name = "DerivativeFixture", BIDSVersion = "1.8.0"),
  file.path(deriv_fixture, "dataset_description.json"),
  auto_unbox = TRUE
)

dir.create(file.path(deriv_fixture, "sub-01", "func"), recursive = TRUE)
file.create(file.path(
  deriv_fixture, "sub-01", "func", "sub-01_task-rest_run-01_bold.nii.gz"
))
readr::write_tsv(
  tibble::tibble(onset = 0, duration = 1, trial_type = "go"),
  file.path(
    deriv_fixture, "sub-01", "func", "sub-01_task-rest_run-01_events.tsv"
  )
)

fmriprep_root <- file.path(deriv_fixture, "derivatives", "fmriprep")
dir.create(file.path(fmriprep_root, "sub-01", "func"), recursive = TRUE)
jsonlite::write_json(
  list(Name = "fmriprep", BIDSVersion = "1.8.0", DatasetType = "derivative"),
  file.path(fmriprep_root, "dataset_description.json"),
  auto_unbox = TRUE
)
file.create(file.path(
  fmriprep_root,
  "sub-01", "func",
  "sub-01_task-rest_run-01_space-MNI_desc-preproc_bold.nii.gz"
))
readr::write_tsv(
  tibble::tibble(
    CSF = c(0.10, 0.20, 0.30, 0.20),
    WhiteMatter = c(0.40, 0.50, 0.60, 0.50),
    GlobalSignal = c(1.00, 1.10, 1.20, 1.10),
    FramewiseDisplacement = c(0.01, 0.02, 0.03, 0.02)
  ),
  file.path(
    fmriprep_root,
    "sub-01", "func",
    "sub-01_task-rest_run-01_desc-confounds_timeseries.tsv"
  )
)

qsiprep_root <- file.path(deriv_fixture, "derivatives", "qsiprep")
dir.create(file.path(qsiprep_root, "sub-01", "anat"), recursive = TRUE)
jsonlite::write_json(
  list(Name = "qsiprep", BIDSVersion = "1.8.0", DatasetType = "derivative"),
  file.path(qsiprep_root, "dataset_description.json"),
  auto_unbox = TRUE
)
file.create(file.path(
  qsiprep_root, "sub-01", "anat", "sub-01_space-MNI_desc-preproc_T1w.nii.gz"
))

proj <- bids_project(deriv_fixture, derivatives = "auto")
```

When a BIDS project already has derivatives, the first job is usually not data
loading. It is figuring out which pipeline produced which files, whether the
expected confounds are present, and whether each run is analysis-ready.

This vignette uses a tiny local project so the full workflow is runnable
without downloading example data. For raw-project discovery, start with
`vignette("quickstart")`.

## Where are the derivatives?

Use `plot_bids()` to get a fast overview of the raw project plus attached
derivative folders.

```{r overview-plot, fig.width=9, fig.height=7}
plot_bids(proj, interactive = FALSE)
```

For code, `derivative_pipelines()` is the clearest inspection entry point.

```{r list-pipelines}
pipes <- derivative_pipelines(proj)
pipes

stopifnot(
  nrow(pipes) == 2L,
  all(c("fmriprep", "qsiprep") %in% pipes$pipeline)
)
```

The fixture has two derivative roots, but only one of them contains functional
preprocessed BOLD data. That is the pipeline you want to target in the next
query.

## How do you pull one derivative run?

Use `query_files()` when you want explicit scope and pipeline control.

```{r query-preproc-run}
prep_bold <- query_files(
  proj,
  regex = "bold\\.nii\\.gz$",
  scope = "derivatives",
  pipeline = "fmriprep",
  desc = "preproc",
  return = "tibble"
)

prep_bold[, c("path", "scope", "pipeline")]

stopifnot(
  nrow(prep_bold) == 1L,
  identical(prep_bold$scope[[1]], "derivatives"),
  identical(prep_bold$pipeline[[1]], "fmriprep")
)
```

That query stays readable because each filter answers one question: which file
suffix, which scope, and which pipeline.

## How do you read and check confounds?

`read_confounds()` reads the derivative table and returns a tibble that stays
indexed by subject, task, run, and session.

```{r read-confounds}
confounds_flat <- read_confounds(
  proj,
  subid = "01",
  task = "rest",
  nest = FALSE
)

confounds_flat

stopifnot(
  nrow(confounds_flat) == 4L,
  all(is.finite(confounds_flat$FramewiseDisplacement)),
  max(confounds_flat$FramewiseDisplacement) < 0.05
)
```

If you want a compact diagnostic view, ask for principal components and plot
them.

```{r plot-confounds}
confounds_pca <- read_confounds(
  proj,
  subid = "01",
  task = "rest",
  npcs = 2
)

plot(confounds_pca, view = "aggregate")
```

The plot is most useful as a quick failure check: if the confounds are missing,
degenerate, or wildly scaled, the PCA summary becomes obviously suspicious.

## How do you verify run-level coverage?

`variables_table()` pulls scans, events, and confounds into one run-level tibble
without forcing you to manage separate joins.

```{r variables-coverage}
run_variables <- variables_table(
  proj,
  scope = "all",
  pipeline = "fmriprep"
)

run_variables[, c(".subid", ".task", ".run", "n_scans", "n_events", "n_confound_rows")]

stopifnot(
  run_variables$n_scans[[1]] == 1L,
  run_variables$n_events[[1]] == 1L,
  run_variables$n_confound_rows[[1]] == 4L
)
```

That table is a practical checkpoint before model fitting because it shows
whether each run has the pieces you expect.

## How do you turn that into a report?

`bids_report()` wraps the same run-level coverage into a compact text report.

```{r print-report}
report <- bids_report(
  proj,
  scope = "all",
  pipeline = "fmriprep"
)

report
```

## Next Steps

Use `vignette("quickstart")` for raw BIDS inspection and `vignette("mock-bids")`
when you need a fully local project for tests, demos, or packaging workflows.

```{r cleanup, include = FALSE}
unlink(deriv_fixture, recursive = TRUE, force = TRUE)
```
