---
title: Create Mock BIDS Projects for Tests and Examples
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
header-includes:
  - '<script src="albers.js"></script>'
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js

vignette: |
  %\VignetteIndexEntry{Create Mock BIDS Projects for Tests and Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```



```{r theme-setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
}

suppressPackageStartupMessages({
  library(bidser)
  library(dplyr)
  library(tidyr)
  library(tibble)
})
```

If you are writing tests, teaching examples, or demonstrating archive
workflows, you usually do not want a downloaded dataset. You want a tiny BIDS
project that lives on disk, has predictable contents, and can include both raw
and derivative files.

`create_mock_bids()` is the entry point for that job.

## What does a small mock project look like?

Start by defining participants and a file inventory.

```{r define-layout}
participants <- tibble(
  participant_id = c("01", "02")
)

file_structure <- tribble(
  ~subid, ~datatype, ~task, ~run, ~suffix, ~fmriprep,
  "01", "func", "rest", "01", "bold.nii.gz", FALSE,
  "01", "func", "rest", "01", "events.tsv", FALSE,
  "01", "func", "rest", "01", "bold.nii.gz", TRUE,
  "01", "func", "rest", "01", "desc-confounds_timeseries.tsv", TRUE,
  "02", "func", "rest", "01", "bold.nii.gz", FALSE,
  "02", "func", "rest", "01", "events.tsv", FALSE
)
```

Now add the tabular payloads you want written into event and confound files.

```{r define-payloads}
event_data <- list(
  "sub-01/func/sub-01_task-rest_run-01_events.tsv" = tibble(
    onset = c(0, 12),
    duration = c(1, 1),
    trial_type = c("cue", "target")
  ),
  "sub-02/func/sub-02_task-rest_run-01_events.tsv" = tibble(
    onset = c(0, 10),
    duration = c(1, 1),
    trial_type = c("cue", "target")
  )
)

confound_data <- list(
  "derivatives/mockprep/sub-01/func/sub-01_task-rest_run-01_desc-confounds_timeseries.tsv" = tibble(
    CSF = c(0.10, 0.20, 0.30, 0.20),
    WhiteMatter = c(0.30, 0.40, 0.50, 0.40),
    GlobalSignal = c(1.00, 1.10, 1.20, 1.10),
    FramewiseDisplacement = c(0.01, 0.02, 0.03, 0.02)
  )
)
```

Create the project on disk so you can query it and archive it later.

```{r create-project}
mock_dir <- tempfile("bidser-mock-")

mock_proj <- create_mock_bids(
  project_name = "RestMock",
  participants = participants,
  file_structure = file_structure,
  event_data = event_data,
  confound_data = confound_data,
  create_stub = TRUE,
  stub_path = mock_dir,
  prep_dir = "derivatives/mockprep"
)

mock_proj

stopifnot(file.exists(file.path(mock_dir, "participants.tsv")))
```

You can inspect the resulting layout immediately with `plot_bids()`.

```{r plot-project}
plot_bids(mock_proj, interactive = FALSE)
```

## How do you query raw and derivative files?

Because `query_files()` works on mock projects too, you can test selection logic
before you point the same code at a real dataset.

```{r query-scope}
raw_runs <- query_files(
  mock_proj,
  regex = "bold\\.nii\\.gz$",
  scope = "raw",
  return = "tibble"
)

derivative_runs <- query_files(
  mock_proj,
  regex = "bold\\.nii\\.gz$",
  scope = "derivatives",
  return = "tibble"
)

raw_runs[, c("path", "scope")]
derivative_runs[, c("path", "scope", "pipeline")]

stopifnot(
  nrow(raw_runs) == 2L,
  nrow(derivative_runs) == 1L
)
```

## How do you inspect injected events and confounds?

The mock project stores real tabular data, so the higher-level readers work the
same way they do on a real project.

```{r inspect-events}
subject_events <- read_events(mock_proj, subid = "01") %>%
  unnest(cols = data)

subject_events

stopifnot(
  nrow(subject_events) == 2L,
  all(subject_events$duration > 0)
)
```

```{r inspect-confounds}
mock_confounds <- read_confounds(
  mock_proj,
  subid = "01",
  nest = FALSE
)

mock_confounds

stopifnot(
  nrow(mock_confounds) == 4L,
  all(is.finite(mock_confounds$GlobalSignal)),
  max(mock_confounds$FramewiseDisplacement) < 0.05
)
```

That makes mock projects useful for both file-level tests and downstream table
checks.

## How do you pack the project for sharing?

`pack_bids()` creates an archive while replacing imaging files with zero-byte
stubs. That keeps the layout and tabular metadata intact without shipping full
images.

```{r pack-project}
archive_file <- tempfile(fileext = ".tar.gz")
packed <- pack_bids(mock_proj, output_file = archive_file, verbose = FALSE)
archive_contents <- list_pack_bids(packed, verbose = FALSE)

archive_contents %>%
  select(file, type, is_stub) %>%
  head()

archive_contents %>%
  summarise(
    n_files = n(),
    n_stubs = sum(is_stub),
    n_tsv = sum(type == "tsv")
  )

stopifnot(
  file.exists(packed),
  any(archive_contents$is_stub)
)
```

## Next Steps

Use `vignette("quickstart")` for reader-facing BIDS discovery and
`vignette("derivatives")` when you want to move from file inventories to
pipeline-aware derivative summaries.

```{r cleanup, include = FALSE}
if (exists("archive_file")) {
  unlink(archive_file, force = TRUE)
}
if (exists("mock_dir")) {
  unlink(mock_dir, recursive = TRUE, force = TRUE)
}
```
