---
title: "Getting Started with bidser"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with bidser}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction to bidser

`bidser` is an R package designed for working with neuroimaging data organized according to the [Brain Imaging Data Structure (BIDS)](https://bids.neuroimaging.io/) standard. BIDS is a specification that describes how to organize and name neuroimaging and behavioral data, making datasets more accessible, shareable, and easier to analyze.

### What is BIDS?

BIDS organizes data into a hierarchical folder structure with standardized naming conventions:

- **Subjects** are identified by folders named `sub-XX`
- **Sessions** (optional) are identified by folders named `ses-XX`
- **Data types** are organized into modality-specific folders (`anat`, `func`, `dwi`, etc.)
- **Files** follow specific naming patterns that encode metadata (subject, session, task, run, etc.)

### What does bidser do?

`bidser` provides tools to:

- **Query and filter** files based on BIDS metadata (subject, task, run, etc.)
- **Read event files** that describe experimental paradigms
- **Work with fMRIPrep derivatives** for preprocessed data
- **Navigate complex BIDS hierarchies** without manually constructing file paths

Let's explore these capabilities using a real BIDS dataset.

## Loading a BIDS Dataset

We'll use the `ds001` dataset from the BIDS examples, which contains data from a "Balloon Analog Risk Task" experiment with 16 subjects.

```{r echo=FALSE}
library(bidser)
```

```{r setup}
library(bidser)
library(tibble)
library(dplyr)
library(gluedown)

# Download example BIDS dataset
tryCatch({
  ds001_path <- get_example_bids_dataset("ds001")
  proj <- bids_project(ds001_path)
  
  print(proj)
}, error = function(e) {
  message("This vignette requires an internet connection to download example data.")
  message("Error: ", e$message)
  knitr::knit_exit()
})
```

The `bids_project` object provides a high-level interface to the dataset. We can see it contains 16 subjects with both anatomical and functional data.

## Basic Dataset Queries

### Dataset Structure

Let's explore the basic structure of this dataset:

```{r}
# Check if the dataset has multiple sessions per subject
sessions(proj)

# Get all participant IDs
participants(proj)

# What tasks are included?
tasks(proj)

# Get a summary of the dataset
bids_summary(proj)
```

### Finding Files by Type

bidser provides several ways to find files. Let's start with the most common neuroimaging file types:

```{r}
# Find all anatomical T1-weighted images
t1w_files <- search_files(proj, regex = "T1w\\.nii", full_path = FALSE)
print(paste("Found", length(t1w_files), "T1w images"))
head(t1w_files)

# Find all functional BOLD scans
bold_files <- func_scans(proj, full_path = FALSE)
print(paste("Found", length(bold_files), "functional scans"))
head(bold_files)
```

### Filtering by Subject and Task

One of bidser's key strengths is filtering data by BIDS metadata:

```{r}
# Get functional scans for specific subjects
sub01_scans <- func_scans(proj, subid = "01")
sub02_scans <- func_scans(proj, subid = "02")

print(paste("Subject 01 has", length(sub01_scans), "scans"))
print(paste("Subject 02 has", length(sub02_scans), "scans"))

# Filter by task (ds001 only has one task, but this shows the syntax)
task_scans <- func_scans(proj, task = "balloonanalogrisktask")
print(paste("Found", length(task_scans), "scans for the balloon task"))

# Combine filters: specific subject AND task
sub01_task_scans <- func_scans(proj, subid = "01", task = "balloonanalogrisktask")
print(paste("Subject 01 has", length(sub01_task_scans), "balloon task scans"))
```

### Working with Multiple Subjects

You can use regular expressions to select multiple subjects at once:

```{r}
# Get scans for subjects 01, 02, and 03
first_three_scans <- func_scans(proj, subid = "0[123]")
print(paste("First 3 subjects have", length(first_three_scans), "scans total"))

# Get scans for all subjects (equivalent to default)
all_scans <- func_scans(proj, subid = ".*")
print(paste("All subjects have", length(all_scans), "scans total"))
```

## Working with Event Files

Event files describe the experimental paradigm - when stimuli were presented, what responses occurred, etc. This is crucial for task-based fMRI analysis.

```{r}
# Find all event files
event_file_paths <- event_files(proj)
print(paste("Found", length(event_file_paths), "event files"))

# Read event data into a nested data frame
events_data <- suppressMessages(read_events(proj))
head(events_data)
```

Let's explore the event data structure:

```{r}
# Look at events for the first subject
library(tidyr)

first_subject_events <- suppressMessages(
  events_data %>% 
    filter(.subid == "01") %>% 
    unnest(cols = c(data))
)

print("Event structure for subject 01:")
print(head(first_subject_events))

# What columns are in the event data?
print("Event file columns:")
print(names(first_subject_events))
```

### Analyzing Event Data

Let's do some basic exploration of the experimental design:

```{r}
# How many trials per subject?
trial_counts <- suppressMessages(
  events_data %>%
    unnest(cols = c(data)) %>%
    group_by(.subid) %>%
    summarise(n_trials = n(), .groups = "drop")
)

print("Trials per subject:")
print(trial_counts)

# What trial types are there?
if ("trial_type" %in% names(first_subject_events)) {
  trial_types <- first_subject_events %>%
    count(trial_type, name = "frequency")
  print("Trial types in the experiment:")
  print(trial_types)
}
```

## Working with Individual Subjects

The `bids_subject()` function provides a convenient interface for working with data from a single subject. It returns a lightweight object with helper functions that automatically filter data for that subject.

```{r}
# Create a subject-specific interface for subject 01
subject_01 <- bids_subject(proj, "01")

# The subject object provides several helper functions:
# Get all functional scans for this subject
sub01_scans <- subject_01$scans()
print(paste("Subject 01 has", length(sub01_scans), "functional scans"))

# Get event files for this subject
sub01_events <- subject_01$events()
print(paste("Subject 01 has", length(sub01_events), "event files"))

# Read event data for this subject
sub01_event_data <- suppressMessages(subject_01$events())
print("Event data structure for subject 01:")
print(sub01_event_data)
```

This approach is particularly useful when you're doing subject-level analyses:

```{r}
# Example: Analyze data for multiple subjects using the subject interface
subjects_to_analyze <- c("01", "02", "03")

for (subj_id in subjects_to_analyze) {
  # Create subject interface
  subj <- bids_subject(proj, subj_id)
  
  # Get subject-specific data
  scans <- subj$scans()
  events <- subj$events()
  
  cat(sprintf("Subject %s: %d scans, %d event files\n", 
              subj_id, length(scans), length(events)))
}
```

The subject interface makes it easy to write analysis pipelines that iterate over subjects without manually constructing filters:

```{r}
# Compare trial counts across subjects using the subject interface
subject_trial_summary <- tibble(
  subject = character(),
  n_trials = numeric(),
  n_scans = numeric()
)

for (subj_id in participants(proj)[1:3]) {  # Just first 3 subjects for demo
  subj <- bids_subject(proj, subj_id)
  
  # Count trials
  event_data <- suppressMessages(subj$events())
  n_trials <- if (nrow(event_data) > 0) {
    event_data %>% 
      unnest(cols = c(data)) %>% 
      nrow()
  } else {
    0
  }
  
  # Count scans
  n_scans <- length(subj$scans())
  
  # Add to summary
  subject_trial_summary <- bind_rows(
    subject_trial_summary,
    tibble(subject = subj_id, n_trials = n_trials, n_scans = n_scans)
  )
}

print("Trial and scan summary by subject:")
print(subject_trial_summary)
```

## Advanced Querying

### Custom File Searches

The `search_files()` function is very flexible for custom queries:

```{r}
# Find all JSON sidecar files
json_files <- search_files(proj, regex = "\\.json$")
print(paste("Found", length(json_files), "JSON files"))

# Find files for specific runs
run1_files <- search_files(proj, regex = "bold", run = "01")
print(paste("Found", length(run1_files), "files from run 01"))

# Complex pattern matching
# Find T1w files for subjects 01-05
t1w_subset <- search_files(proj, regex = "T1w", subid = "0[1-5]")
print(paste("Found", length(t1w_subset), "T1w files for subjects 01-05"))
```

### Getting Full File Paths

Sometimes you need the complete file paths for analysis:

```{r}
# Get full paths to functional scans for analysis
full_paths <- func_scans(proj, subid = "01", full_path = TRUE)
print("Full paths to subject 01's functional scans:")
print(full_paths)

# Check that files actually exist
if (length(full_paths) > 0) {
  files_exist <- file.exists(full_paths)
  print(paste("All files exist:", all(files_exist)))
}
```

## Next Steps

This quickstart covered the basic functionality of bidser for querying BIDS datasets. For more advanced usage, see:

- **fMRIPrep integration**: Working with preprocessed derivatives
- **Data loading**: Reading neuroimaging data with `neurobase` or `RNifti`
- **Confound regression**: Using physiological and motion regressors
- **Group analysis**: Combining data across subjects efficiently

## Reading files produced by FMRIPrep

If you have processed a dataset with FMRIPrep, `bidser` can be used to read in many of the resultant derivative files. If a project has an FMRIPrep derivatives folder, then we can read in the BIDS hierarchy plus derivatives as follows:

```{r}
# Try to get a derivatives dataset - note this may not be available
tryCatch({
  # Some BIDS examples may have derivatives, or we can demonstrate with empty structure
  deriv_path <- get_example_bids_dataset("ds000001-fmriprep")
  proj_deriv <- bids_project(deriv_path, fmriprep=TRUE)
  
  print(proj_deriv)
  
  # Now we can access various derivative files with convenience functions.
  # For example, to read in "preproc" scans we can use the `preproc_scans` function.
  pscans <- preproc_scans(proj_deriv)
  if (!is.null(pscans) && length(pscans) > 0) {
    print(head(as.character(pscans)))
  } else {
    message("No preprocessed scans found in this example dataset")
  }
  
  # Clean up derivatives dataset
  unlink(deriv_path, recursive=TRUE)
  
}, error = function(e) {
  message("Derivatives example not available: ", e$message)
  message("This is normal - not all BIDS examples include fMRIPrep derivatives.")
})
```

```{r cleanup, include=FALSE}
# Clean up the downloaded dataset
if (exists("ds001_path")) {
  unlink(ds001_path, recursive=TRUE)
}
```
